---
title: "SexChromosomeDegenerization"
output: html_document
date: "2024-03-13"
---


```{r setup, include=FALSE}
library(ggplot2)
library(data.table)
library(tidyverse)
library(gridExtra)
library(ggrepel)
library(ggridges)
```


## Script purpose: examine sex chromosome versus other chromosomes in males vs females


## TE analyses

This is looking at chromosome 1 to start with. Are there differences between male and female assembly in TE content? And is this driving the Fst/Pi patterns we see?

```{R chr 1 TEs}
#chr lengths
fem_chrs <- read.table("intermediate_data/female_chrs.tsv", sep = "\t", header = TRUE)
male_chrs <- read.table("intermediate_data/male_chrs.tsv", sep = "\t", header = TRUE)
# lady flies
fem <- read.table("data/female.summary.chr1.tsv", sep = "\t", header = TRUE)
fem$sex <- "Female"
male <- read.table("data/male.summary.chr1.tsv", sep = "\t", header = TRUE)
male$sex <- "Male"

# merge
TEs <- rbind(fem, male)

# remove everything
TEs_fams <- TEs %>% filter(Family != "EVERYTHING_TE")

# barplots
require(scales)
TE_bars <- ggplot(TEs_fams, aes(x=reorder(Family, -Total_Bp), y=Total_Bp))+
  geom_bar(stat='identity')+
  facet_wrap(~sex) +
  coord_flip() + 
  ylab("Total base pairs") +
  xlab("TE family") +
  theme_bw() + scale_y_continuous(labels = comma, breaks = seq(0, 7500000, by = 1250000), limits =c(0,7500000)) # limits = c(0, 5000), breaks = c(0,1250,2500,3750,5000), 

TE_bars

# donut chart of total 
hsize <- 3

# add total length of chr 1:
fem_len <- c("Not_TE", "All_elements", "NA", "NA", 0, "NA", (fem_chrs$Contig_length[1] - sum(as.numeric(fem$Total_Bp))), 0, "female")
fem <- rbind(fem, fem_len)

male_len <- c("Not_TE", "All_elements", "NA", "NA", 0, "NA", (male_chrs$Contig_length[1] - sum(as.numeric(male$Total_Bp))), 0, "male")
male <- rbind(male, male_len)

fem <- fem %>% 
  mutate(x = hsize)

# make numeric
fem$Total_Bp <- as.numeric(fem$Total_Bp)
# drop all TE row
fem <- fem %>% filter(Family != "EVERYTHING_TE")
# add percent
fem$percent <- round(((fem$Total_Bp/fem_chrs$Contig_length[1])*100),2)


fem_pie <- ggplot(fem, aes(x = hsize, y = percent, fill = reorder(Family, percent))) +
  geom_col(color = "black") +
  geom_label_repel(aes(label = paste0(percent, "%")),
                   size = 4.5, nudge_x = 2, show.legend = FALSE) +
  coord_polar(theta = "y") +
 # scale_fill_brewer(palette = "viridis") +
  xlim(c(0.2, hsize + 0.5)) +
  theme(panel.background = element_rect(fill = "white"),
        panel.grid = element_blank(),
        axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank(),
        legend.title=element_blank()) + 
  ggtitle("Female")


fem_pie

male <- male %>% 
  mutate(x = hsize)

# make numeric
male$Total_Bp <- as.numeric(male$Total_Bp)
# drop all TE row
male <- male %>% filter(Family != "EVERYTHING_TE")
male$percent <- round(((male$Total_Bp/male_chrs$Contig_length[1])*100),2)


male_pie <- ggplot(male, aes(x = hsize, y = percent, fill = reorder(Family, percent))) +
  geom_col(color = "black") +
  geom_label_repel(aes(label = paste0(percent, "%")),
                   size = 4.5, nudge_x = 2, show.legend = FALSE) +
  coord_polar(theta = "y") +
 # scale_fill_brewer(palette = "viridis") +
  xlim(c(0.2, hsize + 0.5)) +
  theme(panel.background = element_rect(fill = "white"),
        panel.grid = element_blank(),
        axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank(),
        legend.title=element_blank()) + 
  ggtitle("Male")

male_pie

TEs_fams1 <- TEs_fams
```

Make a merged figure:

```{r chr1 TE merged}
grid.arrange(TE_bars, fem_pie, male_pie, nrow = 2, ncol = 2, #heights = c(1,2),
             layout_matrix = rbind(c(1, 1),
                        c(2,3)))

plot <- arrangeGrob(TE_bars, fem_pie, male_pie, nrow = 2, ncol = 2, #heights = c(1,2),
             layout_matrix = rbind(c(1, 1),
                        c(2,3)))
ggsave("figures/Chr1_TEs.png", plot, width = 9, height = 7, dpi = 600) #  width = 7.5, height = 12,

```

## CHR 2 TEs

```{R chr 2 TEs}
# lady flies
fem <- read.table("data/female.summary.chr2.tsv", sep = "\t", header = TRUE)
fem$sex <- "Female"
male <- read.table("data/male.summary.chr2.tsv", sep = "\t", header = TRUE)
male$sex <- "Male"

# merge
TEs <- rbind(fem, male)

# remove everything
TEs_fams <- TEs %>% filter(Family != "EVERYTHING_TE")

# barplots
require(scales)
TE_bars <- ggplot(TEs_fams, aes(x=reorder(Family, -Total_Bp), y=Total_Bp))+
  geom_bar(stat='identity')+
  facet_wrap(~sex) +
  coord_flip() + 
  ylab("Total base pairs") +
  xlab("TE family") +
  theme_bw() + scale_y_continuous(labels = comma, breaks = seq(0, 5000000, by = 1250000), limits =c(0,5500000)) # limits = c(0, 5000), breaks = c(0,1250,2500,3750,5000), 

TE_bars

# donut chart of total 
hsize <- 3

# add total length of chr 1:
fem_len <- c("Not_TE", "All_elements", "NA", "NA", 0, "NA", (fem_chrs$Contig_length[2] - sum(as.numeric(fem$Total_Bp))), 0, "female")
fem <- rbind(fem, fem_len)

male_len <- c("Not_TE", "All_elements", "NA", "NA", 0, "NA", (male_chrs$Contig_length[2] - sum(as.numeric(male$Total_Bp))), 0, "male")
male <- rbind(male, male_len)

fem <- fem %>% 
  mutate(x = hsize)

# make numeric
fem$Total_Bp <- as.numeric(fem$Total_Bp)
# drop all TE row
fem <- fem %>% filter(Family != "EVERYTHING_TE")
# add percent
fem$percent <- round(((fem$Total_Bp/fem_chrs$Contig_length[2])*100),2)


fem_pie <- ggplot(fem, aes(x = hsize, y = percent, fill = reorder(Family, percent))) +
  geom_col(color = "black") +
  geom_label_repel(aes(label = paste0(percent, "%")),
                   size = 4.5, nudge_x = 2, show.legend = FALSE) +
  coord_polar(theta = "y") +
 # scale_fill_brewer(palette = "viridis") +
  xlim(c(0.2, hsize + 0.5)) +
  theme(panel.background = element_rect(fill = "white"),
        panel.grid = element_blank(),
        axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank(),
        legend.title=element_blank()) + 
  ggtitle("Female")


fem_pie

male <- male %>% 
  mutate(x = hsize)

# make numeric
male$Total_Bp <- as.numeric(male$Total_Bp)
# drop all TE row
male <- male %>% filter(Family != "EVERYTHING_TE")
male$percent <- round(((male$Total_Bp/male_chrs$Contig_length[2])*100),2)


male_pie <- ggplot(male, aes(x = hsize, y = percent, fill = reorder(Family, percent))) +
  geom_col(color = "black") +
  geom_label_repel(aes(label = paste0(percent, "%")),
                   size = 4.5, nudge_x = 2, show.legend = FALSE) +
  coord_polar(theta = "y") +
 # scale_fill_brewer(palette = "viridis") +
  xlim(c(0.2, hsize + 0.5)) +
  theme(panel.background = element_rect(fill = "white"),
        panel.grid = element_blank(),
        axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank(),
        legend.title=element_blank()) + 
  ggtitle("Male")

male_pie

## merge figures
grid.arrange(TE_bars, fem_pie, male_pie, nrow = 2, ncol = 2, #heights = c(1,2),
             layout_matrix = rbind(c(1, 1),
                        c(2,3)))

plot <- arrangeGrob(TE_bars, fem_pie, male_pie, nrow = 2, ncol = 2, #heights = c(1,2),
             layout_matrix = rbind(c(1, 1),
                        c(2,3)))
ggsave("figures/Chr2_TEs.png", plot, width = 9, height = 7, dpi = 600) #  width = 7.5, height = 12,


TEs_fams2 <- TEs_fams
```

## CHR 3 TEs

```{R chr 3 TEs}
# lady flies
fem <- read.table("data/female.summary.chr3.tsv", sep = "\t", header = TRUE)
fem$sex <- "Female"
male <- read.table("data/male.summary.chr3.tsv", sep = "\t", header = TRUE)
male$sex <- "Male"

# merge
TEs <- rbind(fem, male)

# remove everything
TEs_fams <- TEs %>% filter(Family != "EVERYTHING_TE")

# barplots
require(scales)
TE_bars <- ggplot(TEs_fams, aes(x=reorder(Family, -Total_Bp), y=Total_Bp))+
  geom_bar(stat='identity')+
  facet_wrap(~sex) +
  coord_flip() + 
  ylab("Total base pairs") +
  xlab("TE family") +
  theme_bw() + scale_y_continuous(labels = comma, breaks = seq(0, 5000000, by = 1250000), limits =c(0,5500000)) # limits = c(0, 5000), breaks = c(0,1250,2500,3750,5000), 

TE_bars

# donut chart of total 
hsize <- 3

# add total length of chr 1:
fem_len <- c("Not_TE", "All_elements", "NA", "NA", 0, "NA", (fem_chrs$Contig_length[3] - sum(as.numeric(fem$Total_Bp))), 0, "female")
fem <- rbind(fem, fem_len)

male_len <- c("Not_TE", "All_elements", "NA", "NA", 0, "NA", (male_chrs$Contig_length[3] - sum(as.numeric(male$Total_Bp))), 0, "male")
male <- rbind(male, male_len)

fem <- fem %>% 
  mutate(x = hsize)

# make numeric
fem$Total_Bp <- as.numeric(fem$Total_Bp)

# drop all TE row
fem <- fem %>% filter(Family != "EVERYTHING_TE")
# add percent
fem$percent <- round(((fem$Total_Bp/fem_chrs$Contig_length[3])*100),2)


fem_pie <- ggplot(fem, aes(x = hsize, y = percent, fill = reorder(Family, percent))) +
  geom_col(color = "black") +
  geom_label_repel(aes(label = paste0(percent, "%")),
                   size = 4.5, nudge_x = 2, show.legend = FALSE) +
  coord_polar(theta = "y") +
 # scale_fill_brewer(palette = "viridis") +
  xlim(c(0.2, hsize + 0.5)) +
  theme(panel.background = element_rect(fill = "white"),
        panel.grid = element_blank(),
        axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank(),
        legend.title=element_blank()) + 
  ggtitle("Female")


fem_pie

male <- male %>% 
  mutate(x = hsize)

# make numeric
male$Total_Bp <- as.numeric(male$Total_Bp)

# drop all TE row
male <- male %>% filter(Family != "EVERYTHING_TE")
male$percent <- round(((male$Total_Bp/male_chrs$Contig_length[3])*100),2)


male_pie <- ggplot(male, aes(x = hsize, y = percent, fill = reorder(Family, percent))) +
  geom_col(color = "black") +
  geom_label_repel(aes(label = paste0(percent, "%")),
                   size = 4.5, nudge_x = 2, show.legend = FALSE) +
  coord_polar(theta = "y") +
 # scale_fill_brewer(palette = "viridis") +
  xlim(c(0.2, hsize + 0.5)) +
  theme(panel.background = element_rect(fill = "white"),
        panel.grid = element_blank(),
        axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank(),
        legend.title=element_blank()) + 
  ggtitle("Male")

male_pie

## merge figures
grid.arrange(TE_bars, fem_pie, male_pie, nrow = 2, ncol = 2, #heights = c(1,2),
             layout_matrix = rbind(c(1, 1),
                        c(2,3)))

plot <- arrangeGrob(TE_bars, fem_pie, male_pie, nrow = 2, ncol = 2, #heights = c(1,2),
             layout_matrix = rbind(c(1, 1),
                        c(2,3)))
ggsave("figures/Chr3_TEs.png", plot, width = 9, height = 7, dpi = 600) #  width = 7.5, height = 12,


TEs_fams3 <- TEs_fams

```


## CHR 4 TEs

```{R chr 4 TEs}
# lady flies
fem <- read.table("data/female.summary.chr4.tsv", sep = "\t", header = TRUE)
fem$sex <- "Female"
male <- read.table("data/male.summary.chr4.tsv", sep = "\t", header = TRUE)
male$sex <- "Male"

# merge
TEs <- rbind(fem, male)

# remove everything
TEs_fams <- TEs %>% filter(Family != "EVERYTHING_TE")

# barplots
require(scales)
TE_bars <- ggplot(TEs_fams, aes(x=reorder(Family, -Total_Bp), y=Total_Bp))+
  geom_bar(stat='identity')+
  facet_wrap(~sex) +
  coord_flip() + 
  ylab("Total base pairs") +
  xlab("TE family") +
  theme_bw() + scale_y_continuous(labels = comma, breaks = seq(0, 5000000, by = 1250000), limits =c(0,5500000)) # limits = c(0, 5000), breaks = c(0,1250,2500,3750,5000), 

TE_bars

# donut chart of total 
hsize <- 3

# add total length of chr 1:
fem_len <- c("Not_TE", "All_elements", "NA", "NA", 0, "NA", (fem_chrs$Contig_length[4] - sum(as.numeric(fem$Total_Bp))), 0, "female")
fem <- rbind(fem, fem_len)

male_len <- c("Not_TE", "All_elements", "NA", "NA", 0, "NA", (male_chrs$Contig_length[4] - sum(as.numeric(male$Total_Bp))), 0, "male")
male <- rbind(male, male_len)

fem <- fem %>% 
  mutate(x = hsize)

# make numeric
fem$Total_Bp <- as.numeric(fem$Total_Bp)
# drop all TE row
fem <- fem %>% filter(Family != "EVERYTHING_TE")
# add percent
fem$percent <- round(((fem$Total_Bp/fem_chrs$Contig_length[4])*100),2)


fem_pie <- ggplot(fem, aes(x = hsize, y = percent, fill = reorder(Family, percent))) +
  geom_col(color = "black") +
  geom_label_repel(aes(label = paste0(percent, "%")),
                   size = 4.5, nudge_x = 2, show.legend = FALSE) +
  coord_polar(theta = "y") +
 # scale_fill_brewer(palette = "viridis") +
  xlim(c(0.2, hsize + 0.5)) +
  theme(panel.background = element_rect(fill = "white"),
        panel.grid = element_blank(),
        axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank(),
        legend.title=element_blank()) + 
  ggtitle("Female")


fem_pie

male <- male %>% 
  mutate(x = hsize)

# make numeric
male$Total_Bp <- as.numeric(male$Total_Bp)
# drop all TE row
male <- male %>% filter(Family != "EVERYTHING_TE")
male$percent <- round(((male$Total_Bp/male_chrs$Contig_length[4])*100),2)


male_pie <- ggplot(male, aes(x = hsize, y = percent, fill = reorder(Family, percent))) +
  geom_col(color = "black") +
  geom_label_repel(aes(label = paste0(percent, "%")),
                   size = 4.5, nudge_x = 2, show.legend = FALSE) +
  coord_polar(theta = "y") +
 # scale_fill_brewer(palette = "viridis") +
  xlim(c(0.2, hsize + 0.5)) +
  theme(panel.background = element_rect(fill = "white"),
        panel.grid = element_blank(),
        axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank(),
        legend.title=element_blank()) + 
  ggtitle("Male")

male_pie

## merge figures
grid.arrange(TE_bars, fem_pie, male_pie, nrow = 2, ncol = 2, #heights = c(1,2),
             layout_matrix = rbind(c(1, 1),
                        c(2,3)))

plot <- arrangeGrob(TE_bars, fem_pie, male_pie, nrow = 2, ncol = 2, #heights = c(1,2),
             layout_matrix = rbind(c(1, 1),
                        c(2,3)))
ggsave("figures/Chr4_TEs.png", plot, width = 9, height = 7, dpi = 600) #  width = 7.5, height = 12,

TEs_fams4 <- TEs_fams
```

I'm not generally a fan of stacked barplots, but this feels like a good use of them!

```{r stacked barplots}
# merge dfs
fem_chrs$sex <- "Female"
male_chrs$sex <- "Male"

all_chrs <- rbind(fem_chrs, male_chrs)
all_chrs <- all_chrs[,c(1,2,5)]## need co
colnames(all_chrs) <- c("chr", "chr_length", "sex")

# add total TE content for each
TEs_fams1$chr <- 1
TEs_fams2$chr <- 2
TEs_fams3$chr <- 3
TEs_fams4$chr <- 4

# merge
all_TEs_fams <- rbind(TEs_fams1, TEs_fams2, TEs_fams3, TEs_fams4)

#### tidy tidy tidy
# get TE lengths by chromosome per sex
test <- all_TEs_fams %>% group_by(chr, sex) %>% summarise(TE_BP = sum(Total_Bp))
test <- data.frame(test)
# add chr lengths by sex

test2 <- left_join(test,all_chrs, by = c("chr","sex"))

test2$DNA_BP <- test2$chr_length - test2$TE_BP

# need to spread the data to plot
test3 <- test2 %>% gather(type, "length", c(TE_BP,DNA_BP))
test3$type <- gsub(pattern = "_BP", replacement = "", test3$type)

## stacked barplot
ggplot(test3, aes(fill=type, y=length, x=sex)) + 
    geom_bar(position="stack", stat="identity") +
  facet_wrap(~chr) + 
  theme_bw() + 
  ylab("Number of bases") + 
  xlab("Sex") + 
  theme(legend.title= element_blank()) +
  scale_fill_manual(values = c("gray80", "black")) + theme(text=element_text(size=23))
```


TE density: Note this is only the total number of plotted repeats, and not the total content. 

```{R te density plot}
male_TEs <- read.table("data/male_TEpositions.txt",sep="\t",header=F)
colnames(male_TEs) <- c("chr","pos")
male_TEs$sex <- "male"

female_TEs <- read.table("data/female_TEpositions.txt",sep="\t",header=F)
colnames(female_TEs) <- c("chr","pos")
female_TEs$sex <- "female"

# merge
TEs <- rbind(male_TEs, female_TEs)

plottedTEs <- ggplot(TEs) + geom_density(aes(x=pos, color = sex, fill = sex),binwidth=1000000,alpha=.5) + facet_wrap(~chr,ncol=2) + xlab("Genomic position (bins 1 Mb)") + ylab("Number of repeats") + theme_bw()
plottedTEs

plottedTEs2 <- ggplot(TEs, aes(x=pos, y = chr, color = sex, fill = sex))  +  geom_density_ridges(bandwidth=1000000,alpha=.5) +
  theme_ridges() +  xlab("Genomic position") + ylab("Number of repeats") + theme_bw() + ggtitle("B)")
plottedTEs2
```




TE lengths by chr

```{r lengths}

# import data
fem_reps <- fread("data/female_te_lengths.txt")
m_reps <- fread("data/male_te_lengths.txt")
# Rename columns
colnames(fem_reps) <- c("chr", "repeat_begin", "repeat_end", "repeat_type")
colnames(m_reps) <- c("chr", "repeat_begin", "repeat_end", "repeat_type")
# add sex
fem_reps$sex <- "female"
m_reps$sex <- "male"

#merge
reps <- rbind(fem_reps, m_reps)
# calculate length of repeat
reps <- dplyr::mutate(reps, length = repeat_end - repeat_begin)

# summariiiiiieeeees
reps_sum <- reps %>% group_by(repeat_type, chr) %>% summarise(mean_length = mean(length), sd = sd(length), number_of_repeats = n(), minimum_length = min(length), maximum_length = max(length), total_bp = sum(length), Gbp = sum(length)/1000000000)

# total genomic content in repeats here
print(paste0("Total repeat content in the genome from these major families (in Gbp): ", sum(reps_sum$Gbp)))

# save these summary data
#write.csv(reps_sum, "results/RepeatSummaryStats.csv", row.names = FALSE)
#write.csv(reps_sum, "results4supplemental/RepeatSummaryStats.csv", row.names = FALSE)

# drop repeats with very little data (here, < 100Kbp )

reps_sum2 <- reps_sum %>% filter(total_bp > 100000)


# Plot just the summary data!
  
  

sum_plot <- ggplot(reps_sum, aes(x = repeat_type, y = mean_length)) +
  geom_point() + 
  geom_errorbar(aes(ymin = mean_length - sd, ymax = mean_length + sd)) +
  ylab("Length (bp)") +
  xlab("Repeat type") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, color = "black"))
sum_plot


# Now pull out only repeats that are one of types with >100Mbp of repeats

reps_subset <- reps %>% filter(repeat_type %in% reps_sum2$repeat_type)


# violin plot of this...
reps_plot <- ggplot(reps_subset, aes(x = repeat_type, y = length, fill = sex, color = sex)) +
  geom_point(alpha = 0.05, position = "jitter", size = 0.1, pch = ".") +  geom_violin(size = 1, trim = TRUE) + facet_wrap(~chr,ncol=2) +
  ylab("Length (bp)") +
  xlab("Repeat type") + 
  scale_y_log10(labels = scales::comma, breaks = c(10, 100, 1000, 10000, 100000)) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, color = "black")) + ggtitle("C)")
reps_plot

```


Synteny.

```{r synteny}
library(pafr)

ali <- read_paf("scaffolded_y2x.paf")
ali
dotplot(ali, order_by="qstart", dashes = FALSE, ylab = "X chromosome", xlab = "Y chromosome") + coord_flip() + theme_bw() + theme(text=element_text(size=19))


ggsave("figures/XY_synteny.test.png", width = 6, height = 4.5, dpi = 300)

synt <- dotplot(ali, order_by="qstart", dashes = FALSE, ylab = "X chromosome", xlab = "Y chromosome") + coord_flip() + theme_bw() + theme(text=element_text(size=12)) + labs(title = "A)")


```
Make a big plot...


```{R merge te plot}
grid.arrange(synt, plottedTEs2, reps_plot, nrow = 2, ncol = 2, #heights = c(1,2),
             layout_matrix = rbind(c(1, 2),
                        c(3,3)))

TE_plot <- arrangeGrob(synt, plottedTEs2, reps_plot, nrow = 2, ncol = 2, #heights = c(1,2),
             layout_matrix = rbind(c(1, 2),
                        c(3,3)))
ggsave("figures/Fig1_XY_TEs.png", TE_plot, width = 9, height = 10, dpi = 600)


```
